draft

### 2025-05-25 draft.tex

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

@@

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy \citep{pallaSquidpyScalableFramework2022}. Being stand-alone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy. Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts. Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.

To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data. This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization. We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform \citep{CosMxSMIMouse2025}. Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA'). Morphological features were extracted from these 5-channel images for both datasets. Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy \citep{wolfSCANPYLargescaleSinglecell2018} and Pycytominer \citep{serranoReproducibleImagebasedProfiling2025} respectively. We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig. XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data. Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type. Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included. Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated profiling of microscopy data in short scripts and complex pipelines. The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

>>>>

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}. However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

@@

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these widely used features with proven usefulness as a stand-alone Python package is how easy this makes integration into diverse analytical workflows. To demonstrate this, we have developed a module based on cp\_measure for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E stained or fluorescently labeled.

While not the typical application domain of CellProfiler, the recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, using cp\_measure to featurize such images will yield more interpretable features.

However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing. Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, we used cp\_measure to develop a workflow tailored to such data. For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects. This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores. This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler. Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.

To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}. We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue. These masks were then featurised with the described cp\_measure-based workflow. We first preprocessed the features as described in 
\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.

For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data. A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture. Panel C further reveals that some clusters have distinct cell-type compositions: for example, Cluster XXX is dominated by fibroblasts, whereas adipocytes appear almost exclusively in Cluster XXX.

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_4_placeholder.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}

These results demonstrate that our cp\_measure–powered workflow not only scales to the high cell counts typical of spatial datasets but also produces biologically meaningful, interpretable morphological clusters that align with - and complement - transcriptomic phenotypes. By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.

\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated profiling of microscopy data in short scripts and complex pipelines. The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
